---
title: "03_age_profiling"
output: html_document
autor: "María Bueno Álvez"
date: "2024-12-13"
editor_options: 
  chunk_output_type: console
---

# Set up

```{r}
source("scripts/functions/functions_analyses.R")
source("scripts/functions/functions_utility.R")
source("scripts/functions/functions_visualization.R")
source("scripts/functions/themes_palettes.R")

# Read in data
data_disease <- read_tsv("../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_data_phase1.tsv")
meta_disease <- read_tsv("../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_meta_phase1.tsv")

data_disease_wide <- 
  data_disease |> 
  select(-OlinkID, -LOD) |> 
  pivot_wider(names_from = Assay, 
              values_from = NPX) 
```


# ANOVA

```{r}
# Apply the function to each protein group
data_anova <- 
  data_disease |> 
  select(DAid, Assay, NPX) %>%
  left_join(meta_disease |> 
              select(DAid, Disease, Age, Sex, BMI), by = "DAid") |> 
  mutate(BMI = ifelse(BMI == 0, NA, BMI))

# Test the function on one protein
anova_res <- 
  map_df(unique(data_disease$Assay), function(protein) {
    
    do_anova(df = data_anova, 
             protein = protein)
    
  })

saveRDS(anova_res, savepath_data("ANOVA", "anova_wellness.rds"))
anova_res <- readRDS(savepath_data("ANOVA", "anova_wellness.rds"))

write_excel_csv(anova_res, savepath_data("Supplementary-files", "Data_S5.csv"))

anova_res |> 
  filter(term == "Disease",
         p.value < 0.05) 
```


## Visualize

```{r}
dat <- 
  anova_res |> 
  group_by(Protein) |> 
  filter(term != "Residuals") |> 
  mutate(Total_variance = sum(Eta2_perc)) |> 
  arrange(-Total_variance) 

top_variance <- 
  dat |> 
  filter(Total_variance > 45) |> 
  distinct(Protein) |> 
  pull()

#plot_overall <- 
  dat |> 
  ggplot(aes(x = reorder(Protein, -Total_variance), y = Eta2_perc, fill = term)) +
  geom_col() +
  #geom_vline(xintercept = 40, color = "grey", linetype = "dashed", size = 0.5) +
  theme_hpa(angled = T, axis_x_title = T) +
  scale_fill_manual(values = pal_anova) +
  ylab("Variance explained (%)") +
  xlab("Proteins") 


ggsave(savepath_results("Manuscript-figures", "anova_DA.pdf"), h = 4, w = 13)

# Top proteins
plot_top <- 
  dat |> 
  filter(Protein %in% top_variance) |>
  ggplot(aes(x = reorder(Protein, -Total_variance), y = Eta2_perc, fill = term)) +
  geom_col(show.legend = F) +
  theme_hpa(angled = T) +
  scale_fill_manual(values = pal_anova) +
  ylab("% of variance explained") +
  xlab("")+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_text(size = 8),
        axis.line.x = element_blank())

plot_overall / plot_top

ggsave(savepath_results("Manuscript-figures", "anova_DA.pdf"), h = 6, w = 12)


groups <- 
  dat |> 
  group_by(Protein) |> 
  top_n(1, Eta2_perc) |> 
  select(Protein, top_term = term)

selected_groups <- 
  groups |> 
  filter(top_term != "Disease")
dat |> 
  group_by(Protein) |> 
  top_n(1, Eta2_perc) |> 
  ungroup() |> 
  count(term)

dat |>  
  inner_join(selected_groups) |>  
  ggplot(aes(x = reorder(Protein, -Total_variance), y = Eta2_perc, fill = term)) +  
  geom_col() +  
  geom_vline(xintercept = 40, color = "grey", linetype = "dashed", size = 0.5) +  
  theme_hpa(angled = T) +  
  scale_fill_manual(values = pal_anova) +  
  facet_grid(. ~ top_term, scales = "free", space = "free") +  # Use facet_grid instead of facet_wrap
  ylab("% of variance explained") +  
  xlab("")




#Top protein for age, sex, BMI

dat_gdf15 <- 
  data_disease |> 
  filter(Assay == "GDF15") |> 
  left_join(meta_disease |> 
              select(DAid, Sex, Class), by = "DAid")  
  
order_gdf15 <- 
  dat_gdf15 |> 
  group_by(Class) |>
  summarise(mean = mean(NPX, na.rm = T)) |> 
  arrange(mean)

p1 <- 
  dat_gdf15 |> 
  mutate(Class = factor(Class, levels = order_gdf15$Class)) |>
  ggplot(aes(Class, NPX, color = Class, fill = Class)) +
  geom_quasirandom(alpha = 0.6, show.legend = F) +
  geom_boxplot(fill = NA, outlier.colour = NA, color = "grey20") +
  # geom_text(aes(label = "adj p-value = 1.16e-297")) +
  scale_color_manual(values = pal_class) +
  scale_fill_manual(values = pal_class) +
  theme_hpa(axis_x_title = T) +
  ggtitle("GDF15")  +
  xlab("Disease class")

p2 <- 
  data_disease |> 
  filter(Assay == "CGA") |> 
  left_join(meta_disease |> 
              select(DAid, Sex, Class), by = "DAid") |> 
  filter(!is.na(Sex)) |> 
  mutate(Sex = case_when(Sex == "M" ~ "Male",
                         Sex == "F" ~ "Female"),
         Sex = factor(Sex, levels = c("Male", "Female"))) |> 
  ggplot(aes(Sex, NPX, color = Class, fill = Class)) +
  geom_quasirandom(alpha = 0.6, show.legend = F) +
  geom_boxplot(fill = NA, outlier.colour = NA, color = "grey20") +
 # geom_text(aes(label = "adj p-value = 1.16e-297")) +
  scale_color_manual(values = pal_class) +
  scale_fill_manual(values = pal_class) +
  theme_hpa() +
  ggtitle("CGA") 

p3 <- 
  data_disease |> 
  filter(Assay == "EDA2R") |> 
  left_join(meta_disease |> 
              select(DAid, Age, Class), by = "DAid") |> 
  filter(!is.na(Age)) |> 
  ggplot(aes(Age, NPX, color = Class)) +
  geom_point(show.legend = F) +
  geom_smooth(color = "grey20") +
  scale_color_manual(values = pal_class) +
  theme_hpa() +
  ggtitle("EDA2R")
  
p4 <- 
  data_disease |> 
  filter(Assay == "LEP") |> 
  left_join(meta_disease |> 
              select(DAid, BMI, Class), by = "DAid") |> 
  filter(!is.na(BMI),
         BMI != 0) |> 
  ggplot(aes(BMI, NPX, color = Class)) +
  geom_point(show.legend = F) +
  geom_smooth(color = "grey20") +
  scale_color_manual(values = pal_class) +
  theme_hpa() +
  ggtitle("LEP")

p3 | p4 | p1 | p2

ggsave(savepath_results("Manuscript-figures", "age_sex_bmi_prots.pdf"), h = 3, w = 12)


dat |>  
  group_by(term) |>  
  top_n(15, Eta2_perc) |>  
  filter(term != "Disease") |>  
  mutate(term = factor(term, levels = c("Sex", "Age", "BMI")),
         Protein = reorder_within(Protein, Eta2_perc, term)) |>  # Reorder within each term
  ggplot(aes(Protein, Eta2_perc, color = term)) +  # Change fill to color for lines
  geom_segment(aes(xend = Protein, yend = 0), size = 1) +  # Lollipop stem
  geom_point(size = 3, aes(fill = term), shape = 21) +  # Lollipop head
  facet_wrap(~term, scales = "free", ncol = 1) +  
  scale_color_manual(values = pal_anova) +  # Color for the stems
  scale_fill_manual(values = pal_anova) +  # Fill for the points
  scale_x_reordered() +  # Apply reordering to x-axis
  xlab("") +  
  theme_hpa(angled = T) +  
  coord_flip() +
  theme(panel.spacing = unit(1.5, "lines"))

ggsave(savepath("age_sex_bmi_top_prots_lollipop.pdf"), w = 3.5, h = 11)



```

# Machine learning

## Generate initial splits

### Age

```{r}
age_data <- 
  data_disease_wide |> 
  left_join(meta_disease |> 
              select(DAid, Age), by = "DAid") |> 
  filter(!is.na(Age),
         Age >=2)

age_ml_split <- 
  generate_split(data = age_data, 
                 proportion = 0.8,
                 variable_stratify = "Age")

#saveRDS(age_ml_split, "server-scripts/data/age_ml_split.rds")
age_ml_split <- readRDS("server-scripts/data/age_ml_split.rds")

```

### Sex

```{r}
sex_data <- 
  data_disease_wide |> 
  left_join(meta_disease |> 
              select(DAid, Sex), by = "DAid") |> 
  filter(!is.na(Sex))

sex_ml_split <- 
  generate_split(data = sex_data, 
                 proportion = 0.8,
                 variable_stratify = "Sex")

#saveRDS(sex_ml_split, "server-scripts/data/sex_ml_split.rds")
sex_ml_split <- readRDS("server-scripts/data/sex_ml_split.rds")

```

### BMI

```{r}
bmi_data <- 
  data_disease_wide |> 
  left_join(meta_disease |> 
              select(DAid, BMI), by = "DAid") |> 
  filter(!is.na(BMI),
         BMI != 0)

bmi_ml_split <- 
  generate_split(data = bmi_data, 
                 proportion = 0.8,
                 variable_stratify = "BMI")

#saveRDS(bmi_ml_split, "server-scripts/data/bmi_ml_split.rds")
bmi_ml_split <- readRDS("server-scripts/data/bmi_ml_split.rds")
```


## Load results from seed analyes (HPC)

```{r}
# Protein importances
protein_importance_age <- read_tsv("server-scripts/parsed-results/importances_age.tsv")
protein_importance_sex <- read_tsv("server-scripts/parsed-results/importances_sex.tsv")
protein_importance_bmi <- read_tsv("server-scripts/parsed-results/importances_bmi.tsv")

# Predictions
predictions_age <- read_tsv("server-scripts/parsed-results/predictions_age.tsv")
predictions_sex <- read_tsv("server-scripts/parsed-results/predictions_sex.tsv")
predictions_bmi <- read_tsv("server-scripts/parsed-results/predictions_bmi.tsv")
```

```{r}
protein_importance_age |> 
  filter(abs(Importance) > 0) |> 
  count(Seed) |> 
  arrange(n)


protein_importance_sex |> 
  filter(abs(Importance) > 0) |> 
  count(Seed) |> 
  arrange(-n)


protein_importance_bmi |> 
  filter(abs(Importance) > 0) |> 
  count(Seed) |> 
  arrange(n)

```

## Results summary 

```{r}
p1 <- plot_importance_frequency(protein_importance_sex, "Sex")
p2 <- plot_importance_frequency(protein_importance_age, "Age")
p3 <- plot_importance_frequency(protein_importance_bmi, "BMI")

plot_n_feature <- p1 | p2 | p3
```


## Top 10

```{r}
top_sex <- 
  plot_top_proteins_seed(protein_importance =  protein_importance_sex,
                       plot_color = pal_anova[["Sex"]],
                       n = 10) 
top_age <- 
  plot_top_proteins_seed(protein_importance =  protein_importance_age,
                       plot_color = pal_anova[["Age"]],
                       n = 10)
top_bmi <- 
  plot_top_proteins_seed(protein_importance =  protein_importance_bmi,
                        plot_color = pal_anova[["BMI"]],
                       n = 10)


plot_top_proteins <- top_sex | top_age | top_bmi
```


## Performance

```{r}
# # Within-testing set - discrete(sex)
# ml_age$performance |>
#   filter(.metric == "rsq") |> 
#   mutate(Type = "Age") |> 
#   bind_rows(ml_bmi$performance |>
#   filter(.metric == "rsq") |> 
#   mutate(Type = "BMI")) |> 
#   bind_rows(ml_sex$performance |>
#   mutate(Type = "Sex")) |>
#   mutate(Type = factor(Type, levels = c("Sex", "Age", "BMI"))) |> 
#   ggplot(aes(Type, .estimate, color = Type, fill = Type)) +
#   geom_quasirandom() +
#   geom_boxplot(alpha = 0.5,
#                color = "grey40",
#                outlier.color = NA) +
#   scale_color_manual(values = pal_age_sex_bmi) +
#   scale_fill_manual(values = pal_age_sex_bmi) + 
#   theme_hpa() +
#   ggtitle("Performance across seeds")
# 
# ggsave(savepath("performance_age_sex_bmi.png"), h = 8, w = 8)
```


# Build final models (25 proteins)

## Sex model

```{r}
models_sex <- 
  map_df(seq(5, 200, by = 5), function(n) {
    
    cat(paste0("\n Fitting model using ", n, " proteins"))
    
    sex_proteins <- 
      protein_importance_sex |> 
      filter(Importance > 0) |> 
      group_by(Variable) |> 
      summarise(Avg_importance = mean(Importance)) |> 
      arrange(-Avg_importance) |> 
      head(n)
    
    sex_train <- 
      sex_ml_split$data_train  |> 
      select(DAid, sex_proteins$Variable, Sex)
    
    sex_test <- 
      sex_ml_split$data_test |> 
      select(DAid, sex_proteins$Variable, Sex)
    
    sex_prediction <-  
      discrete_prediction(split_train = sex_train,
                          split_test = sex_test,
                          variable_predict = "Sex",
                          variable_case = "F")  
  
  tibble(n_proteins = n, 
         performance = sex_prediction$performance |> filter(.metric == "roc_auc") |> pull(.estimate) |> round(2))
  
  
})

p_sex <- 
  models_sex |> 
  ggplot(aes(n_proteins, performance)) +
  geom_point(color = pal_anova[["Sex"]]) +
  geom_line(color = pal_anova[["Sex"]]) +
  geom_vline(xintercept = 15, lty = "dashed", color = "grey") +
  theme_hpa() +
  xlab("Number of proteins") +
  ylab("Performance (R2)")

```


## Age model

```{r}
models_age <- 
  map_df(seq(5, 200, by = 5), function(n) {
  
  
  cat(paste0("\n Fitting model using ", n, " proteins"))
    
  age_proteins <- 
    protein_importance_age |> 
    filter(Importance > 0) |> 
    group_by(Variable) |> 
    summarise(Avg_importance = mean(Importance)) |> 
    arrange(-Avg_importance) |> 
    head(n)
  
  age_train <- 
    age_ml_split$data_train |> 
    select(DAid, age_proteins$Variable, Age)
  
  
  age_test <- 
    age_ml_split$data_test |> 
    select(DAid, age_proteins$Variable, Age)
  
  age_prediction <-  
    continuous_prediction(split_train = age_train,
                          split_test = age_test,
                          variable_predict = "Age")  
  
  tibble(n_proteins = n, 
         performance = age_prediction$performance |> filter(.metric == "rsq") |> pull(.estimate) |> round(2))
  
  
})



p_age <- 
  models_age |> 
  ggplot(aes(n_proteins, performance)) +
  geom_point(color = pal_anova[["Age"]]) +
  geom_line(color = pal_anova[["Age"]]) +
  geom_vline(xintercept = 50, lty = "dashed", color = "grey") +
  theme_hpa() +
  xlab("Number of proteins") +
  ylab("Performance (R2)")
```

### Specific 

```{r}
age_proteins <- 
    protein_importance_age |> 
    filter(Importance > 0) |> 
    group_by(Variable) |> 
    summarise(Avg_importance = mean(Importance)) |> 
    arrange(-Avg_importance) |> 
    head(50)
  
plot_top_proteins_seed(protein_importance =  protein_importance_age,
                        plot_color = pal_anova[["Age"]],
                       n = 50)

ggsave(savepath_results("Manuscript-figures", "Top_50_age.pdf"), h = 10, w = 4)


  age_train <- 
    age_ml_split$data_train |> 
    select(DAid, age_proteins$Variable, Age)
  
  
  age_test <- 
    age_ml_split$data_test |> 
    select(DAid, age_proteins$Variable, Age)
  
  age_prediction <-  
    continuous_prediction(split_train = age_train,
                          split_test = age_test,
                          variable_predict = "Age")  
  
R2 <- 
  age_prediction$performance |> 
  filter(.metric == "rsq") |> 
  pull(.estimate) |> 
  round(2)

#predictions_age <- 
age_prediction$predictions |> 
  mutate(DAid = age_ml_split$data_test$DAid) |> 
  left_join(meta_disease, by = "DAid") |> 
  mutate(offset = Variable - .pred) |> 
  ggplot(aes(Variable, .pred, color = offset)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, lty = "dashed") +
  geom_text(aes(label = paste0("R2 = ", R2), 
                x = 75,
                y = 0.25), inherit.aes = F) +
  scale_colour_gradient2() +
  theme_hpa() +
  ggtitle("Age")

ggsave(savepath_results("Manuscript-figures", "Fig3e_age_prediction.pdf"), h = 6, w = 6)


age_prediction$predictions |> 
  mutate(DAid = age_ml_split$data_test$DAid) |> 
  left_join(meta_disease, by = "DAid") |> 
  mutate(offset = Variable - .pred,
         Class = factor(Class, levels = class_order)) |> 
  ggplot(aes(Variable, .pred, color = Class)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, lty = "dashed") +
 # geom_text(aes(label = paste0("R2 = ", R2), 
  #              x = 75,
   #             y = 0.25), inherit.aes = F) +
  scale_color_manual(values = pal_class) +
  facet_wrap(~Class, nrow = 2) +
  theme_hpa() +
  ggtitle("Age")

ggsave(savepath_results("Manuscript-figures", "Class_age_prediction.pdf"), h = 6, w = 9)
```

### Proteins

```{r}

age_carrasco_zanini_2024 <- 
  age_sex_bmi_carrasco_2024 |> 
  filter(`Top variable (olink)` == "age") |> 
  distinct(Assay)|> 
  mutate(Study = "Carrasco - 2024") |> 
  filter(Assay %in% unique(data_disease$Assay))

age_sun_2023 <- 
  age_sex_bmi_sun_2023 |> 
  filter(`Age log10(p)` > 2.3) |> 
  filter(Assay %in% unique(data_disease$Assay)) |> 
  top_n(150, `Age log10(p)`) |> 
  distinct(Assay)|> 
  mutate(Study = "Sun - 2023")
  
age_eldjarn_2023 <- 
  age_sex_eldjarn_2023 |> 
  filter(gene_name %in% unique(data_disease$Assay)) |> 
  mutate(pval_age_af = as.double(pval_sex_af)) |> 
  filter(pval_age_af < 0.05) |> 
 # top_n(150, pval_sex_af) |> 
  distinct(Assay = gene_name) |> 
  mutate(Study = "Eldjarn - 2023")

age_argentieri_2023 <- 
  age_clock |> 
  distinct(Assay = Gene_name) |> 
  mutate(Study = "Argentieri - 2023")|> 
  filter(Assay %in% unique(data_disease$Assay))


age_argentieri_subset_2023 <- 
  age20 |> 
  distinct(Assay = Gene_name)|> 
  mutate(Study = "Argentieri subset - 2023")|> 
  filter(Assay %in% unique(data_disease$Assay))

plot_proteins <- 
  plot_top_proteins_seed(protein_importance =  protein_importance_age,
                        plot_color = pal_anova[["Age"]],
                       n = 50)

plot_study <- 
  age_proteins |> 
  rename(Assay = Variable) |> 
  mutate(Carrasco = ifelse(Assay %in% age_carrasco_zanini_2024$Assay, "Yes", "No"),
         Sun = ifelse(Assay %in% age_sun_2023$Assay, "Yes", "No"),
         Eldjar = ifelse(Assay %in% age_eldjarn_2023$Assay, "Yes", "No"),
         Argentieri = ifelse(Assay %in% age_argentieri_2023$Assay, "Yes", "No"),
         Argentieri_subset = ifelse(Assay %in% age_argentieri_subset_2023$Assay, "Yes", "No")) |> 
  pivot_longer(cols = c(Carrasco, Sun, Eldjar, Argentieri, Argentieri_subset), 
               names_to = "Study", values_to = "In_study") |>
  mutate(Assay = factor(Assay, levels = rev(age_proteins$Variable)),
         Study = factor(Study, levels = c("Eldjar", "Sun", "Argentieri", "Argentieri_subset", "Carrasco"))) |> 
  ggplot(aes(Study, Assay, fill = In_study)) +
  geom_tile(color = "white") +
  coord_fixed() +
  scale_fill_manual(values = c("grey90", pal_anova[["Age"]])) +
  theme_hpa(angled = T, axis_y = F) +
  xlab("")

plot_proteins | plot_study
ggsave(savepath_results("Manuscript-figures", "proteins_age_studies.pdf"), h = 10, w = 6)

```

## BMI model

```{r}
models_bmi <- 
  map_df(seq(5, 200, by = 5), function(n) {
  
  
  cat(paste0("\n Fitting model using ", n, " proteins"))
    
  bmi_proteins <- 
    protein_importance_bmi |> 
    filter(Importance > 0) |> 
    group_by(Variable) |> 
    summarise(Avg_importance = mean(Importance)) |> 
    arrange(-Avg_importance) |> 
    head(n)
  
  bmi_train <- 
    bmi_ml_split$data_train |> 
    select(DAid, bmi_proteins$Variable, BMI)
  
  
  bmi_test <- 
    bmi_ml_split$data_test |> 
    select(DAid, bmi_proteins$Variable, BMI)
  
  bmi_prediction <-  
    continuous_prediction(split_train = bmi_train,
                          split_test = bmi_test,
                          variable_predict = "BMI")  
  
  tibble(n_proteins = n, 
         performance = bmi_prediction$performance |> filter(.metric == "rsq") |> pull(.estimate) |> round(2))
  
  
})

p_bmi <- 
  models_bmi |> 
  ggplot(aes(n_proteins, performance)) +
  geom_point(color = pal_anova[["BMI"]]) +
  geom_line(color = pal_anova[["BMI"]]) +
  geom_vline(xintercept = 40, lty = "dashed", color = "grey") +
  theme_hpa() +
  xlab("Number of proteins") +
  ylab("Performance (R2)")
```

## Combine

```{r}
plot_performances <- p_sex + p_age + p_bmi

ggsave(savepath_results("Manuscript-figures", "performance_age_sex_bmi.png"), h = 5 , w = 14)
```


# Compare with ANOVA

```{r}
ml_res_summary <- 
  protein_importance_sex |> 
  group_by(Variable) |> 
  summarise(mean_importance = mean(abs(Importance))) |> 
  rename(Protein = Variable) |> 
  mutate(Variable = "Sex") |> 
  bind_rows(protein_importance_age |> 
              group_by(Variable) |> 
              summarise(mean_importance = mean(abs(Importance)))|> 
              rename(Protein = Variable) |> 
              mutate(Variable = "Age")) |> 
  bind_rows(protein_importance_bmi |> 
              group_by(Variable) |> 
              summarise(mean_importance = mean(abs(Importance)))|> 
              rename(Protein = Variable) |> 
              mutate(Variable = "BMI"))

anova_res_summary <- 
  anova_res |> 
  filter(!term %in% c("Disease", "Residuals")) |> 
  select(Protein, Variable = term, Eta2_perc)

dat_anova_comparison <- 
  ml_res_summary |> 
  left_join(anova_res_summary, by = c("Protein", "Variable")) |> 
  mutate(Variable = factor(Variable, levels = c("Sex", "Age", "BMI"))) 

p1 <- 
  dat_anova_comparison |> 
  filter(Variable == "Sex") |> 
  ggplot(aes(mean_importance, Eta2_perc)) +
  geom_point(color = pal_anova[["Sex"]]) +
  geom_text_repel(aes(label = Protein), color = pal_anova[["Sex"]]) +
  theme_hpa() +
  xlab("avg_importance")

p2 <- 
  dat_anova_comparison |> 
  filter(Variable == "Age") |> 
  ggplot(aes(mean_importance, Eta2_perc)) +
  geom_point(color = pal_anova[["Age"]]) +
  geom_text_repel(aes(label = Protein), color = pal_anova[["Age"]]) +
  theme_hpa() +
  xlab("avg_importance")

labels_bmi <- 
  dat_anova_comparison |> 
  filter(Variable == "BMI") |> 
  top_n(2, Eta2_perc) 
p3 <- 
  dat_anova_comparison |> 
  filter(Variable == "BMI") |> 
  ggplot(aes(mean_importance, Eta2_perc)) +
  geom_point(color = pal_anova[["BMI"]]) +
  geom_text_repel(data = labels_bmi, aes(label = Protein), color = pal_anova[["BMI"]]) +
  theme_hpa() +
  xlab("avg_importance")

plot_anova_comparison <-  p1 | p2 | p3
```

# Combine into suppl

```{r}
plot_n_feature /
plot_top_proteins /
plot_performances /
plot_anova_comparison 

ggsave(savepath_results("Manuscript-figures", "Suppl_age_sex_BMI.pdf"), h = 13, w = 10)
```



# Check age proteins

## Import age proteins

```{r}
age_clock <- 
  import_df("data/UKB/proteomics_clock.xlsx")
age20 <- 
  import_df("data/UKB/protage_20_clock.xlsx")
```

## Top 10 in disease atlas

```{r}
data_disease |> 
  filter(Assay %in% age20$Gene_name) |> 
  left_join(meta_disease |> 
              select(DAid, Age, Class), by = "DAid") |> 
  mutate(Assay = factor(Assay, levels = age20$Gene_name)) |> 
  ggplot(aes(Age, NPX, color = Class)) +
  geom_point(size = 0.5) +
#  geom_smooth() +
  facet_wrap(~Assay, scales = "free_y") +
  scale_color_manual(values = pal_class) +
  theme_hpa()

ggsave(savepath("age20_disease_atlas.png"), h = 6, w = 8)
```

# Examples

```{r}
p1 <- 
  data_disease |> 
  filter(Assay == "CXCL17") |> 
  left_join(meta_disease |> 
              select(DAid, Age, Class), by = "DAid") |> 
  ggplot(aes(Age, NPX, color = Class)) +
  geom_point(size = 0.5, show.legend = F) +
  geom_smooth(color = "black") +
  scale_color_manual(values = pal_class) +
  theme_hpa() 
  
p2 <- 
  data_disease |> 
  filter(Assay == "COL9A1") |> 
  left_join(meta_disease |> 
              select(DAid, Age, Class), by = "DAid") |> 
  ggplot(aes(Age, NPX, color = Class)) +
  geom_point(size = 0.5, show.legend = F) +
  geom_smooth(color = "black") +
  scale_color_manual(values = pal_class) +
  theme_hpa()

p1 / p2
ggsave(savepath_results("Manuscript-figures", "Fig3c-d_examples_age.pdf"), h = 8, w = 4)

p2
ggsave(savepath_results("Manuscript-figures", "col9a1_age.pdf"), h = 4, w = 4)
```


## Overlap between lists

```{r}

y <- list("Age clock" = age_clock$Gene_name,
          "Age clock - 20" = age20$Gene_name,
          "ML" = age_proteins$Variable)

plot(euler(y, shape = "ellipse"), quantities = T, fills = c("#D0E4EE","grey50", pal_anova["Age"])) |> as.ggplot() 

ggsave(savepath_results("Manuscript-figures", "overlap_age.pdf"), h = 3, w = 3)


top_proteins <- 
    ml_age$protein_importance|> 
    group_by(term) |> 
    summarise(abs_estimate = mean(abs(estimate)),
              n = n_distinct(seed)) |> 
    arrange(-abs_estimate) |> 
    head(50)

age_proteins |> 
    mutate(Variable = factor(Variable, levels = rev(age_proteins$Variable)),
         Class = case_when(Variable %in% age20$Gene_name ~ "Argentieri_clock_20",
                           Variable %in% age_clock$Gene_name & !Variable %in% age20$Gene_name ~ "Argentieri_clock_extended",
                           T ~ "ML")) |> 
  ggplot(aes(fct_reorder(Variable, Avg_importance), Avg_importance, color = Class, fill = Class)) +
  geom_quasirandom(size = 0.5, show.legend = F) +
  geom_boxplot(alpha = 0.5, outlier.color = NA, show.legend = F) +
  scale_color_manual(values = c("Argentieri_clock_extended" = "#D0E4EE","Argentieri_clock_20" = "grey50", "ML" = pal_anova[["Age"]])) +
  scale_fill_manual(values = c("Argentieri_clock_extended" = "#D0E4EE","Argentieri_clock_20" = "grey50", "ML" = pal_anova[["Age"]])) +
  coord_flip() +
  xlab("")  +
  ylab("") +
  theme_hpa() 

ggsave(savepath("top50_age.pdf"), h = 10, w = 4)
ggsave(savepath("top50_age.pdf"), h = 10, w = 8)
```


```{r}
top_proteins <- 
  ml_age$protein_importance |> 
  group_by(term) |> 
  summarise(abs_estimate = mean(abs(estimate))) |> 
  arrange(-abs_estimate) |> 
  head(50)

ml_age$protein_importance  |> 
  filter(term %in% top_proteins$term) |> 
  mutate(term = factor(term, levels = rev(top_proteins$term)),
         Age_clock = ifelse(term %in% age_clock$Gene_name, "Yes", "No")) |> 
  ggplot(aes(term, abs(estimate), color = Age_clock, fill = Age_clock)) +
  geom_quasirandom(size = 0.5) +
  geom_boxplot(alpha = 0.5, outlier.color = NA) +
  coord_flip() +
  scale_color_manual(values = rev(c("#0077b6", "grey"))) +
  scale_fill_manual(values = rev(c("#0077b6", "grey"))) +
  theme_hpa()

ggsave(savepath("age_proteins_clock.pdf"), h = 10, w = 6)



p1 <- 
  plot_boxplot(proteins = c("CXCL17", "LTBP2"),
             type = "age") +
  ggtitle("In age clock") 
p2 <- 
  plot_boxplot(proteins = c("CCL27", "ART3"),
               type = "age") +
  ggtitle("Not in age clock")

p1/p2
ggsave(savepath("age_proteins.png"), h = 10, w = 10)

```

# Age

```{r}
ml_pan_disease  <-
  protein_importance_age |> 
  group_by(Variable) |> 
  filter(Importance > 0) |> 
  summarise(Avg_importance = mean(Importance),
            n = n_distinct(Seed)) |> 
  filter(n > 90) |>
  top_n(150, Avg_importance) |>
  distinct(Assay = Variable) |> 
  mutate(Study = "Current - ML")

age_carrasco_zanini_2024 <- 
  age_sex_bmi_carrasco_2024 |> 
  filter(`Top variable (olink)` == "age") |> 
  distinct(Assay)|> 
  mutate(Study = "Carrasco - 2024") |> 
  filter(Assay %in% unique(data_disease$Assay))

age_sun_2023 <- 
  age_sex_bmi_sun_2023 |> 
  # filter(`Age log10(p)` > 2.3,
  #        `Age Beta` > 0.02) |> 
  filter(Assay %in% unique(data_disease$Assay)) |> 
  top_n(150, `Age log10(p)`) |> 
  distinct(Assay)|> 
  mutate(Study = "Sun - 2023")
  
age_eldjarn_2023 <- 
  age_sex_eldjarn_2023 |> 
  filter(gene_name %in% unique(data_disease$Assay)) |> 
  mutate(pval_age_af = as.double(pval_sex_af)) |> 
  filter(pval_age_af < 0.05) |> 
  top_n(150, pval_sex_af) |> 
  distinct(Assay = gene_name) |> 
  mutate(Study = "Eldjarn - 2023")

age_argentieri_2023 <- 
  age_clock |> 
  distinct(Assay = Gene_name) |> 
  mutate(Study = "Argentieri - 2023")|> 
  filter(Assay %in% unique(data_disease$Assay))


age_argentieri_subset_2023 <- 
  age20 |> 
  distinct(Assay = Gene_name)|> 
  mutate(Study = "Argentieri subset - 2023")|> 
  filter(Assay %in% unique(data_disease$Assay))

combined_age <- 
  ml_pan_disease |> 
  bind_rows(age_carrasco_zanini_2024) |> 
  bind_rows(age_sun_2023) |> 
  bind_rows(age_eldjarn_2023) |> 
  bind_rows(age_argentieri_2023) #|> 
  #bind_rows(age_argentieri_subset_2023)

library(ggupset)
pdf(savepath_results("Manuscript-figures", "upset_age.pdf"), h = 5, w = 10)

combined_age |> 
  group_by(Assay) |> 
  summarise(Study = list(unique(Study))) |> 
  ggplot(aes(x=Study)) +
  geom_bar(fill = pal_anova[["Age"]]) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -1) +
  scale_x_upset(n_intersections = 20) +
  theme_hpa() 
  
dev.off()
 ###
age_sex_bmi_carrasco_2024 <- 
  import_df("data/UKB/age_sex_bmi_Carrasco_2024.xlsx") 

age_sex_bmi_sun_2023 <- 
  import_df("data/UKB/age_sex_bmi_Sun_2023.xlsx") 

age_sex_eldjarn_2023 <- 
  import_df("data/UKB/age_sex_Eldjarn_2023.xlsx") 

sex
age
bmi

```

# Sex

```{r}
sex_ml_pan_disease  <-
  protein_importance_sex |> 
  group_by(Variable) |> 
  filter(Importance > 0) |> 
  summarise(Avg_importance = mean(Importance),
            n = n_distinct(Seed)) |> 
  filter(n > 90) |>
  top_n(150, Avg_importance) |>
  distinct(Assay = Variable) |> 
  mutate(Study = "Current - ML")

sex_carrasco_zanini_2024 <- 
  age_sex_bmi_carrasco_2024 |> 
  filter(`Top variable (olink)` == "sex") |> 
  distinct(Assay)|> 
  mutate(Study = "Carrasco - 2024") |> 
  filter(Assay %in% unique(data_disease$Assay))

sex_sun_2023 <- 
  age_sex_bmi_sun_2023 |> 
  # filter(`Age log10(p)` > 2.3,
  #        `Age Beta` > 0.02) |> 
  filter(Assay %in% unique(data_disease$Assay)) |> 
  top_n(150, `Sex log10(p)`) |> 
  distinct(Assay)|> 
  mutate(Study = "Sun - 2023")
  
sex_eldjarn_2023 <- 
  age_sex_eldjarn_2023 |> 
  filter(gene_name %in% unique(data_disease$Assay)) |> 
  mutate(pval_sex_af = as.double(pval_sex_af)) |> 
  filter(pval_sex_af < 0.05) |> 
  top_n(150, pval_sex_af) |> 
  distinct(Assay = gene_name) |> 
  mutate(Study = "Eldjarn - 2023")

combined_sex <- 
  sex_ml_pan_disease |> 
  bind_rows(sex_carrasco_zanini_2024) |> 
  bind_rows(sex_sun_2023) |> 
  bind_rows(sex_eldjarn_2023) 

pdf(savepath_results("Manuscript-figures", "upset_sex.pdf"), h = 5, w = 5)
combined_sex |> 
  group_by(Assay) |> 
  summarise(Study = list(unique(Study))) |> 
  ggplot(aes(x=Study)) +
  geom_bar(fill = pal_anova[["Sex"]]) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -1) +
  scale_x_upset(n_intersections = 20) +
  theme_hpa()
dev.off()

```

# BMI

```{r}
bmi_ml_pan_disease  <-
  protein_importance_bmi |> 
  group_by(Variable) |> 
  filter(Importance > 0) |> 
  summarise(Avg_importance = mean(Importance),
            n = n_distinct(Seed)) |> 
  filter(n > 90) |>
  top_n(150, Avg_importance) |>
  distinct(Assay = Variable) |> 
  mutate(Study = "Current - ML")

bmi_carrasco_zanini_2024 <- 
  age_sex_bmi_carrasco_2024 |> 
  filter(`Top variable (olink)` == "bmi") |> 
  distinct(Assay)|> 
  mutate(Study = "Carrasco - 2024") |> 
  filter(Assay %in% unique(data_disease$Assay))

bmi_sun_2023 <- 
  age_sex_bmi_sun_2023 |> 
  # filter(`Age log10(p)` > 2.3,
  #        `Age Beta` > 0.02) |> 
  filter(Assay %in% unique(data_disease$Assay)) |> 
  top_n(150, `BMI log10(p)`) |> 
  distinct(Assay)|> 
  mutate(Study = "Sun - 2023")
  

combined_bmi <- 
  sex_ml_pan_disease |> 
  bind_rows(bmi_carrasco_zanini_2024) |> 
  bind_rows(bmi_sun_2023) 

pdf(savepath_results("Manuscript-figures", "upset_bmi.pdf"), h = 5, w = 3)

combined_bmi |> 
  group_by(Assay) |> 
  summarise(Study = list(unique(Study))) |> 
  ggplot(aes(x=Study)) +
  geom_bar(fill = pal_anova[["BMI"]]) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -1) +
  scale_x_upset(n_intersections = 20) +
  theme_hpa()

dev.off()
```

