---
title: "04_disease_profiling_overview"
output: html_document
autor: "María Bueno Álvez"
date: "2024-12-13"
editor_options: 
  chunk_output_type: console
---

# Set up

```{r}
source("scripts/functions/functions_analyses.R")
source("scripts/functions/functions_utility.R")
source("scripts/functions/functions_visualization.R")
source("scripts/functions/themes_palettes.R")

# Read in data
data_disease <-
  read_tsv(
    "../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_data_phase1.tsv"
  )
meta_disease <-
  read_tsv(
    "../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_meta_phase1.tsv"
  )
```

## Disease order

```{r}
disease_class_order <-
  meta_disease |>
  distinct(Class, Disease) |>
  mutate(Class = factor(Class, levels = class_order)) |>
  arrange(Class, Disease) |>
  pull(Disease)

diseases <-
  meta_disease |>
  filter(Class != "Healthy") |>
  distinct(Disease) |>
  mutate(Disease = factor(Disease, levels = disease_class_order)) |>
  arrange(Disease) |>
  pull()

classes <-
  meta_disease |>
  filter(Class != "Healthy") |>
  distinct(Class) |>
  pull()
```

# Export data summaries

Extract and export processed data summaries for supplementary files.

## Proteomics data

```{r}
data_summarised <-
  data_disease |>
  left_join(meta_disease |>
              select(DAid, Disease), by = "DAid") |>
  group_by(Disease, Assay, OlinkID) |>
  # Generate summaries
  summarise(
    mean = mean(NPX),
    median = median (NPX),
    Q1 = quantile(NPX, .25),
    Q3 = quantile(NPX, .75)
  ) |>
  # Round to two decimals
  mutate(across(c(mean, median, Q1, Q3), ~ round(.x, 2)),
         Disease = factor(Disease, levels = disease_class_order)) |>
  arrange(Disease, Assay)

write_excel_csv(data_summarised,
                savepath_data("Supplementary-files", "Data_S12.csv"))
```

## Cohort info

```{r}
cohort_info <- 
  meta_disease |> 
  summarise(
    Class = first(Class),  # Assuming Class is unique per Disease
    Age_range = paste0(round(mean(Age, na.rm = TRUE), 2), 
                       " (", round(min(Age, na.rm = TRUE), 2), 
                       " - ", round(max(Age, na.rm = TRUE), 2), ")"),
    `Males (n)` = sum(Sex == "M", na.rm = TRUE),
    `Females (n)` = sum(Sex == "F", na.rm = TRUE),
    .by = Disease
  ) |> 
  mutate(Disease = factor(Disease, levels = disease_class_order)) |> 
  arrange(Disease)

  
write_excel_csv(cohort_info, savepath_data("Supplementary-files", "Data_S13.csv"))
```

# Disease overview

Generate disease-specific summaries to provide an overview of cohort composition and sample distribution.

## Barplot

```{r}
meta_disease |>
  count(Disease, Class) |>
  mutate(Disease = factor(Disease, levels = rev(disease_class_order))) |>
  ggplot(aes(Disease, n, fill = Class)) +
  geom_col(show.legend = F) +
  coord_flip() +
  scale_fill_manual(values = pal_class) +
  theme_hpa() +
  xlab("") +
  ylab("Number of samples")

ggsave(
  savepath_results("Manuscript-figures", "Fig_cohort_overview.pdf"),
  h = 9,
  w = 8
)
```

## Semicircle number of samples

```{r}
data <-
  data_disease |>
  left_join(meta_disease, by = "DAid") |>
  filter(DAid %in% meta_disease$DAid) |>
  group_by(Disease, Class) |>
  summarise(n = n()) |>
  ungroup() |>
  mutate(
    fraction = n / sum(n),
    Disease = factor(Disease, levels = disease_class_order)
  ) |>
  arrange(Disease)


data$ymax = cumsum(data$fraction)
data$ymin = c(0, head(data$ymax, n = -1))

data |>
  ggplot(aes(
    ymax = ymax,
    ymin = ymin,
    xmax = 4,
    xmin = 3.5,
    fill = Class
  )) +
  geom_rect(show.legend = F) +
  coord_polar(theta = "y") +
  ylim(c(0, 2)) +
  xlim(c(2, 4)) +
  scale_fill_manual(values = pal_class) +
  theme_void()

ggsave(savepath_results("Manuscript-figures", "n_sample_overview.pdf"), h = 5, w = 5)
```

## Age, Sex, BMI distirbution

```{r}
sex_plot <-
  meta_disease |>
  group_by(Class, Disease, Sex) |>
  summarize(n = n_distinct(DAid)) |>
  ungroup() |>
  group_by(Disease) |>
  mutate(total_n = sum(n)) |>
  mutate(Disease = factor(Disease, levels = (disease_class_order))) |>
  ggplot(aes(Disease, n, fill = Sex)) +
  geom_col() +
  geom_text(aes(label = total_n, y = total_n + 60)) +
  scale_fill_manual(values = pal_sex) +
  theme_hpa(angled = T, axis_y = T) +
  coord_flip() +
  theme(legend.position = "top") +
  xlab("") +
  theme(axis.ticks.y = element_blank(),
        axis.line.y = element_blank())

age_plot <-
  meta_disease |>
  mutate(Disease = factor(Disease, levels = (disease_class_order))) |>
  ggplot(aes(
    x = Age,
    y = Disease,
    color = Class,
    fill = Class,
    group = Disease
  )) +
  geom_density_ridges(
    alpha = 0.5,
    quantile_lines = TRUE,
    quantiles = 2,
    show.legend = F
  ) +
  scale_fill_manual(values = pal_class) +
  scale_color_manual(values = pal_class) +
  theme_hpa(angled = T, axis_y = F)

bmi_plot <-
  meta_disease |>
  mutate(BMI = ifelse(BMI == 0, NA, BMI)) |>
  mutate(Disease = factor(Disease, levels = (disease_class_order))) |>
  ggplot(aes(
    x = BMI,
    y = Disease,
    color = Class,
    fill = Class,
    group = Disease
  )) +
  geom_density_ridges(
    alpha = 0.5,
    quantile_lines = TRUE,
    quantiles = 2,
    show.legend = F
  ) +
  scale_fill_manual(values = pal_class) +
  scale_color_manual(values = pal_class) +
  theme_hpa(angled = T, axis_y = F)

sex_plot | age_plot | bmi_plot
ggsave(
  savepath_results("Manuscript-figures", "S_Fig_cohort_overview_ridges.pdf"),
  h = 14,
  w = 12
)
```

## Sample UMAP 

```{r}
umap_disease <- do_umap(data_disease,
                        wide = F,
                        impute = T)

disease_centers <-
  umap_disease |>
  left_join(meta_disease, by = c("Sample" = "DAid")) |>
  group_by(Disease, Class) |>
  summarise(UMAP1 = median(UMAP1),
            UMAP2 = median(UMAP2))

umap_disease |>
  left_join(meta_disease, by = c("Sample" = "DAid")) |>
  ggplot(aes(UMAP2, UMAP1, color = Class)) +
  geom_point(alpha = 0.5, show.legend = F) +
  geom_text(data = disease_centers, aes(label = Disease), show.legend = F) +
  scale_color_manual(values = pal_class) +
  theme_hpa()

ggsave(
  savepath_results("Manuscript-figures", "Fig4_UMAP.pdf"),
  h = 8,
  w = 6
)
```


## Disease correlations

```{r}
cor_disease <-
  data_disease |>
  left_join(meta_disease |>
              select(DAid, Disease), by = "DAid") |>
  group_by(Disease, Assay) |>
  summarise(NPX = mean(NPX)) |>
  pivot_wider(names_from = "Assay", values_from = "NPX") |>
  column_to_rownames("Disease") |>
  scale() |>
  t() |>
  cor(use = "complete.obs", method = "spearman")

annotation <-
  meta_disease |>
  distinct(Disease, Class) |>
  column_to_rownames("Disease")

# Mask the upper triangle
ord = rownames(cor_disease)
hc = hclust(as.dist(1 - cor_disease))
cor_disease = cor_disease[hc$order, hc$order]
cor_disease[upper.tri(cor_disease)] = NA
cor_disease = cor_disease[ord, ord]

cor_disease |>
  pheatmap(
    clustering_method = "ward.D2",
    color = pal_cor,
    cluster_col = hc,
    cluster_row = hc,
    annotation_col = annotation,
    annotation_row = annotation,
    annotation_colors = list("Class" = pal_class),
    na_col = "white"
  ) |>
  as.ggplot()

ggsave(
  savepath_results(
    "Manuscript-figures",
    "Fig1_correlation_diseases_scaled.pdf"
  ),
  h = 12,
  w = 15
)
```

# Case-control examples

Extract representative case–control comparisons to illustrate the power of a pan-disease approach.

## Prepare data

```{r}
ml_data <- 
  data_disease |>
  left_join(meta_disease |> 
              select(DAid, Disease, Class), by = "DAid") |> 
  select(DAid, Disease, Assay, NPX) |> 
  pivot_wider(names_from = Assay,
              values_from = NPX) 

set.seed(213)
disease_split <-
  ml_data |> 
  initial_split(prop = 0.7,  strata = Disease)

disease_train <- training(disease_split)
disease_test <- testing(disease_split)
```

## FGF21

```{r}
# Define current protein and classes to classify against
protein <- "FGF21"

disease <-  "Pancreatic cancer"

selected_bacterial_infections <-
  c(
    "E.coli pyelonephritis",
    "Pneumococcal pneumonia",
    "Staphylococcus aureus bacteremia",
    "Streptococcal soft tissue infection"
  )

res_lr_classes <-
  current_meta <-
  meta_disease |>
  mutate(
    Disease_grouped = case_when(
      Disease == "Pancreatic cancer" ~ "Pancreatic cancer",
      Disease %in%  selected_bacterial_infections ~
        "Selected bacterial infections ",
      Disease == "Healthy" ~ "Healthy",
      T ~ "Other"
    )
  )

current_classes <-
  current_meta |>
  distinct(Disease_grouped) |>
  filter(Disease_grouped != disease) |>
  pull()

# Run logistic regression against all classes
current_res <-
  map(current_classes, function(class) {
    control_diseases <-
      current_meta |>
      filter(Disease_grouped == class) |>
      distinct(Disease) |>
      filter(Disease != disease) |>
      pull()
    
    do_log_reg (
      protein = protein,
      disease_case = disease,
      disease_control = control_diseases,
      disease_train = disease_train,
      disease_test = disease_test
    )
    
  }) |>
  set_names(current_classes)

# Extract ROC results
roc_class_df <-
  map_df(current_classes, function(class) {
    res <-  current_res[[class]]
    
    res$roc_curve |>
      mutate(AUC = res$AUC) |>
      mutate(
        Class = class,
        Disease = disease,
        Protein = protein,
        Combo = paste(disease, protein, sep = "_")
      )
  })

# Plot ROC curves
tiles <-
  tibble(
    specificity = c(0.2, 0.2, 0.2),
    sensitivity = c(0.4, 0.3, 0.2),
    Class = current_classes
  )  |>
  left_join(roc_class_df |>
              distinct(Combo, Class, AUC), by = "Class") |>
  mutate(AUC = round(AUC, 2)) |>
  arrange(-AUC)

current_pal <-
  c(
    "Other" = "darkred",
    "Selected bacterial infections " = pal_class[["Infection"]],
    "Healthy" = pal_class["Healthy"]
  )

tiles <-
  tiles |>
  mutate(Class = factor(Class, levels = tiles$Class))

plot_fgf21 <-
  roc_class_df |>
  mutate(Class = factor(Class, levels = tiles$Class)) |>
  ggplot(aes(
    x = 1 - specificity,
    y = sensitivity,
    group = Class
  )) +
  geom_path(aes(color = Class), size = 1, show.legend = F) +
  geom_segment(aes(
    x = 0,
    y = 0,
    xend = 1,
    yend = 1
  ),
  colour = 'grey',
  linetype = 'dotdash') +
  geom_tile(
    data = tiles,
    aes(fill = Class),
    alpha = 0.7,
    show.legend = F,
    height = 0.1,
    width = 0.1
  ) +
  geom_text(
    data = tiles,
    aes(label = AUC),
    size = 3,
    color = "black",
    show.legend = F
  ) +
  geom_text(
    data = tiles,
    aes(label = Class,
        x = 1 - specificity - 0.2),
    size = 3,
    color = "black",
    show.legend = F
  ) +
  geom_text(
    label = "AUC",
    x = 0.8,
    y = 0.5,
    size = 3,
    inherit.aes = F
  ) +  
  scale_color_manual(values = current_pal) +
  scale_fill_manual(values = current_pal) +
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 1)) +
  theme_hpa() +
  coord_fixed()
```

## GSTA3

```{r}
# Define current protein and classes to classify against
protein <- "GSTA3"

disease <- "Hepatocellular cancer"

selected_liver <-
  c(
    "Alcohol-related liver disease",
    "Chronic liver disease",
    "MASLD",
    "Viral hepatitis related cirrhosis"
  )
res_lr_classes <-
  
  current_meta <-
  meta_disease |>
  mutate(
    Disease_grouped = case_when(
      Disease == "Hepatocellular cancer" ~ "Hepatocellular cancer",
      Disease %in%  selected_liver ~ "Liver diseases",
      Disease == "Healthy" ~ "Healthy",
      T ~ "Other"
    )
  )

current_classes <-
  current_meta |>
  distinct(Disease_grouped) |>
  filter(Disease_grouped != disease) |>
  pull()

# Run logistic regression against all classes
current_res <-
  map(current_classes, function(class) {
    control_diseases <-
      current_meta |>
      filter(Disease_grouped == class) |>
      distinct(Disease) |>
      filter(Disease != disease) |>
      pull()
    
    do_log_reg (
      protein = protein,
      disease_case = disease,
      disease_control = control_diseases,
      disease_train = disease_train,
      disease_test = disease_test
    )
    
  }) |>
  set_names(current_classes)

# Extract ROC results
roc_class_df <-
  map_df(current_classes, function(class) {
    res <-  current_res[[class]]
    
    res$roc_curve |>
      mutate(AUC = res$AUC) |>
      mutate(
        Class = class,
        Disease = disease,
        Protein = protein,
        Combo = paste(disease, protein, sep = "_")
      )
  })

# Plot ROC curves
tiles <-
  tibble(
    specificity = c(0.2, 0.2, 0.2),
    sensitivity = c(0.4, 0.3, 0.2),
    Class = current_classes
  )  |>
  left_join(roc_class_df |>
              distinct(Combo, Class, AUC), by = "Class") |>
  mutate(AUC = round(AUC, 2)) |>
  arrange(-AUC)


current_pal <-
  c(
    "Other" = "darkred",
    "Liver diseases" = "#E5C494",
    "Healthy" = pal_class[["Healthy"]]
  )
tiles <-
  tiles |>
  mutate(Class = factor(Class, levels = tiles$Class))

plot_gsta3 <-
  roc_class_df |>
  mutate(Class = factor(Class, levels = tiles$Class)) |>
  ggplot(aes(
    x = 1 - specificity,
    y = sensitivity,
    group = Class
  )) +
  geom_path(aes(color = Class), size = 1, show.legend = F) +
  geom_segment(aes(
    x = 0,
    y = 0,
    xend = 1,
    yend = 1
  ),
  colour = 'grey',
  linetype = 'dotdash') +
  geom_tile(
    data = tiles,
    aes(fill = Class),
    alpha = 0.7,
    show.legend = F,
    height = 0.1,
    width = 0.1
  ) +
  geom_text(
    data = tiles,
    aes(label = AUC),
    size = 3,
    color = "black",
    show.legend = F
  ) +
  geom_text(
    data = tiles,
    aes(label = Class,
        x = 1 - specificity - 0.2),
    size = 3,
    color = "black",
    show.legend = F
  ) +
  geom_text(
    label = "AUC",
    x = 0.8,
    y = 0.5,
    size = 3,
    inherit.aes = F
  ) +  # Fixed the warning
  scale_color_manual(values = current_pal) +
  scale_fill_manual(values =  current_pal) +
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 1)) +
  theme_hpa() +
  coord_fixed()
```


## CES3

```{r}
# Define current protein and classes to classify against
protein <- "CES3"

diseases <-  c("Rheumatoid arthritis",
             "Systemic lupus erythematosus")

current_meta <-
  meta_disease |>
  mutate(
    Disease_grouped = case_when(
      Disease %in% diseases ~ Disease,
      Disease == "Dengue" ~ "Dengue",
      Disease == "Healthy" ~ "Healthy",
      T ~ "Other"
    )
  )

current_classes <-
  current_meta |>
  distinct(Disease_grouped) |>
  filter(!Disease_grouped %in% diseases) |>
  pull()

# Run logistic regression against all classes
current_res <-
  map(current_classes, function(class) {
    control_diseases <-
      current_meta |>
      filter(Disease_grouped == class) |>
      distinct(Disease) |>
      filter(Disease != disease) |>
      pull()
    
    do_log_reg (
      protein = protein,
      disease_case = diseases,
      disease_control = control_diseases,
      disease_train = disease_train,
      disease_test = disease_test
    )
    
  }) |>
  set_names(current_classes)

# Extract ROC results
roc_class_df <-
  map_df(current_classes, function(class) {
    res <-  current_res[[class]]
    
    res$roc_curve |>
      mutate(AUC = res$AUC) |>
      mutate(
        Class = class,
        Disease = disease,
        Protein = protein,
        Combo = paste(disease, protein, sep = "_")
      )
  })

# Plot ROC curves
tiles <-
  tibble(
    specificity = c(0.2, 0.2, 0.2),
    sensitivity = c(0.4, 0.3, 0.2),
    Class = current_classes
  )  |>
  left_join(roc_class_df |>
              distinct(Combo, Class, AUC), by = "Class") |>
  mutate(AUC = round(AUC, 2)) |>
  arrange(-AUC)


current_pal <-
  c(
    "Other" = "darkred",
    "Dengue" = pal_class[["Infection"]],
    "Healthy" = pal_class["Healthy"]
  )
tiles <-
  tiles |>
  mutate(Class = factor(Class, levels = tiles$Class))

plot_ces3 <-
  roc_class_df |>
  mutate(Class = factor(Class, levels = tiles$Class)) |>
  ggplot(aes(
    x = 1 - specificity,
    y = sensitivity,
    group = Class
  )) +
  geom_path(aes(color = Class), size = 1, show.legend = F) +
  geom_segment(aes(
    x = 0,
    y = 0,
    xend = 1,
    yend = 1
  ),
  colour = 'grey',
  linetype = 'dotdash') +
  geom_tile(
    data = tiles,
    aes(fill = Class),
    alpha = 0.7,
    show.legend = F,
    height = 0.1,
    width = 0.1
  ) +
  geom_text(
    data = tiles,
    aes(label = AUC),
    size = 3,
    color = "black",
    show.legend = F
  ) +
  geom_text(
    data = tiles,
    aes(label = Class,
        x = 1 - specificity - 0.2),
    size = 3,
    color = "black",
    show.legend = F
  ) +
  geom_text(
    label = "AUC",
    x = 0.8,
    y = 0.5,
    size = 3,
    inherit.aes = F
  ) +  # Fixed the warning
  scale_color_manual(values = current_pal) +
  scale_fill_manual(values = current_pal) +
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 1)) +
  theme_hpa() +
  coord_fixed()
```

## Combine 

```{r}
plot_fgf21 / plot_ces3 / plot_gsta3
ggsave(savepath("auc_biomarkers.pdf"), h = 12, w = 3)
```
