---
title: "02_disease_profiling"
output: html_document
autor: "María Bueno Álvez"
date: "2024-12-13"
editor_options: 
  chunk_output_type: console
---

# Set up

```{r}
source("scripts/functions/functions_analyses.R")
source("scripts/functions/functions_utility.R")
source("scripts/functions/functions_visualization.R")
source("scripts/functions/themes_palettes.R")

# Read in data
data_disease <- read_tsv("../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_data_phase1.tsv")
meta_disease <- read_tsv("../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_meta_phase1.tsv")
```

## Disease order

```{r}
disease_class_order <-
  meta_disease |> 
  distinct(Class, Disease) |> 
  mutate(Class = factor(Class, levels = class_order)) |> 
  arrange(Class) |> 
  pull(Disease)

diseases <- 
  meta_disease |> 
  filter(Class != "Healthy") |> 
  distinct(Disease) |> 
  mutate(Disease = factor(Disease, levels = disease_class_order)) |> 
  arrange(Disease) |> 
  pull()

classes <- 
  meta_disease |> 
  filter(Class != "Healthy") |> 
  distinct(Class) |> 
  pull()

# Exclude diseases with more than 50 samples for machine learning
exclude_diseases <- 
  meta_disease |>
  filter(DAid %in% data_disease$DAid) |>
  count(Disease) |> 
  arrange(n) |> 
  filter(n < 48) |> 
  pull(Disease)

include_diseases <- 
  meta_disease |>
  distinct(Disease) |> 
  filter(!Disease %in% c(exclude_diseases, "Healthy")) |> 
  pull(Disease)

disease_class_order_ml <-
  meta_disease |> 
  distinct(Class, Disease) |> 
  filter(Disease %in% include_diseases) |> 
  arrange(Class) |> 
  pull(Disease)
```

## Prepare data

```{r}
ml_data <- 
  data_disease |>
  left_join(meta_disease |> 
              select(DAid, Disease, Class), by = "DAid") |> 
  filter(Disease %in% c(include_diseases, "Healthy")) |> 
  select(DAid, Disease, Assay, NPX) |> 
  pivot_wider(names_from = Assay,
              values_from = NPX) 


# Split data 100 times
seeds <- c(1:100)
random_splits <-
  map(seeds, \(seed) generate_split(data = ml_data,
                                    proportion = 0.7,
                                    variable_stratify = "Disease",
                                    seed = seed))

#saveRDS(random_splits, "server-scripts/data/random_splits.rds")
```

## Read results from cluster 

```{r}
ml_multiclass <- readRDS("server-scripts/parsed-results/multiclass_multiclass_results.rds")

ml_class <- 
  map(c("Cardiovascular", "Metabolic", "Cancer", "Autoimmune", "Infection"), \(class) {
    
    readRDS(paste0("scripts/random/summer/parsed-results/", class, "_multiclass_results.rds"))
    
  })

ml_healthy <- 
  map(c("Cardiovascular", "Metabolic", "Cancer", "Psychiatric", "Autoimmune", "Infection"), \(class) {
    
     readRDS(paste0("scripts/random/summer/parsed-results/binary_results/", class, "_binary_results.rds"))
    
  })

ml_multiclass$predictions


# Importances
importances_all <- read_tsv("server-scripts/parsed-results/importances_disease_all.tsv")
importances_class <- read_tsv("server-scripts/parsed-results/importances_disease_class.tsv")
importances_healthy <- read_tsv("server-scripts/parsed-results/importances_disease_healthy.tsv")

# Performance
predictions_all <- read_tsv("server-scripts/parsed-results/predictions_disease_all.tsv")
predictions_class <- read_tsv("server-scripts/parsed-results/roc_disease_class.tsv")

# ROC
roc_all <- read_tsv("server-scripts/parsed-results/roc_disease_all.tsv")
roc_class <- read_tsv("server-scripts/parsed-results/roc_disease_class.tsv")
roc_healthy <- read_tsv("server-scripts/parsed-results/roc_healthy_seed1.tsv")

```

## Feature importance

```{r}
importances_class |> 
  group_by(class, Seed) |> 
  summarise(n = n_distinct(term)) |> 
  left_join(meta_disease |> 
              distinct(Disease, Class), by = c("class" = "Disease")) |> 
  # filter(class =="Coronary artery calcification",
  #        n > 250)
  
    mutate(class = factor(class, levels = disease_class_order_ml)) |>
  ggplot(aes(class, n, color = Class, group = Seed)) +
  geom_quasirandom() +
  geom_line(color = "grey90") +
 # geom_boxplot(fill = NA, color = "black", outlier.colour = NA) +
  scale_color_manual(values = pal_class) +
  theme_hpa(angled = T)

```


### Extracts importances

```{r}
class_importance <- 
  map_df(ml_class, \(ml_res) {
    
   ml_res$protein_importance
    
  })

healthy_importance <- 
  map_df(ml_healthy, \(ml_res) {
    
   ml_res$protein_importance |> 
      rename(class = Disease)
    
  })

multiclass_importance <- ml_multiclass$protein_importance

# multiclass_importance |> 
#   ungroup() |> 
#   group_by(class) |> 
#   summarise(n = n_distinct(seed)) |> 
#   arrange(n) |> View()

# ranked_glioma_25 <- 
#   multiclass_importance |> 
#   filter(class == "Glioma") |> 
#   count(term) |> 
#   arrange(-n) |> 
#   head(25)
# 
# multiclass_importance |> 
#   filter(class == "Glioma",
#          term %in% ranked_glioma_25$term) |> 
#   plot_top_proteins_seed(plot_color = pal_class[["Cancer"]]) +
#   ggtitle("Glioma")
#   
# ggsave(savepath("glioma_multiclass.png"), h = 5, w = 5)



# ranked_vte_25 <- 
#   multiclass_importance |> 
#   filter(class == "Acute venous thromboembolism") |> 
#   count(term) |> 
#   arrange(-n) |> 
#   head(25)
# 
# multiclass_importance |> 
#   filter(class == "Acute venous thromboembolism",
#          term %in% ranked_vte_25$term) |> 
#   plot_top_proteins_seed(plot_color = pal_class[["Cardiovascular"]]) +
#   ggtitle("Actute venous thromboembolism")
#   
# ggsave(savepath("vte_multiclass.png"), h = 5, w = 5)

```


### Protein importance for example diseases

```{r}
importances_all  |> 
  distinct(class) |> 
  filter(!class %in% include_diseases)

example_diseases <- c("Acute myeloid leukemia",
                      "Influenza",
                      "Acute coronary syndrome",
                      "Rheumatoid arthritis",
                      "Viral hepatitis related cirrhosis", 
                      "Schizophrenia")


importances_all |> 
  mutate(class = ifelse(class == "Scleroderma", "Systemic sclerosis", class)) |> 
  group_by(class, term) |> 
  summarise(avg_estimate = mean(estimate)) |> 
  group_by(class) |> 
  top_n(n = 20, wt = abs(avg_estimate)) |> 
  arrange(class, avg_estimate) |> 
  mutate(Names = paste(class, avg_estimate)) |> 
  left_join(meta_disease |> 
              distinct(Disease, Class), by = c("class" = "Disease")) |> 
  filter(class %in% example_diseases) |> 
  #filter(class != "Healthy") |> 
 # mutate(class = factor(class, levels = disease_class_order[disease_class_order %in% include_diseases])) |> 
  ggplot(aes(fct_reorder(Names, -avg_estimate), avg_estimate, color = Class)) + 
  geom_point(show.legend = F) + 
  geom_segment(aes(x = Names, xend = Names, y = 0, yend = avg_estimate), show.legend = F) + 
  geom_text(aes(label = term), show.legend = F, angle = 90, size = 2) + 
  geom_hline(yintercept = 0) + 
  facet_wrap(~class, scales = "free_x", nrow = 1) + 
  scale_color_manual(values = pal_class) + 
  xlab("") + 
  theme_hpa(axis_x = F)

ggsave(savepath("features-multiclass.pdf"), h = 3, w = 12)

```


### Number important features

```{r}
plot_n_importance(class_importance)
ggsave(savepath("features_class.png"), h = 5, w = 10)

plot_n_importance(multiclass_importance)
ggsave(savepath("features_multiclass.png"), h = 5, w = 10)

plot_n_importance(healthy_importance)
ggsave(savepath("features_binary.png"), h = 5, w = 10)

# Have it faceted


```

### Number of time a feature is selected

```{r}
multiclass_importance |>
  group_by(class, term) |> 
  summarize(n = n_distinct(seed)) |> 
  left_join(resource_meta |> distinct(Disease, Class), by = c("class" = "Disease")) |>
  mutate(class = factor(class, levels = disease_class_order)) |> 
  #filter(n > 25) |> 
  ggplot(aes(class, n, color = Class, fill = Class)) +
  geom_quasirandom(size = 0.5) +
  geom_boxplot(width = 0.5, outlier.color = NA, color = "grey20", alpha = 0.5) +
  scale_color_manual(values = pal_class) + 
  scale_fill_manual(values = pal_class) + 
  theme_hpa(angled = T)

healthy_importance |>
  group_by(class, term) |> 
  summarize(n = n_distinct(seed)) |> 
  left_join(resource_meta |> distinct(Disease, Class), by = c("class" = "Disease")) |>
  mutate(class = factor(class, levels = disease_class_order)) |> 
  #filter(n > 25) |> 
  ggplot(aes(class, n, color = Class, fill = Class)) +
  geom_quasirandom(size = 0.5) +
  geom_boxplot(width = 0.5, outlier.color = NA, color = "grey20", alpha = 0.5) +
  scale_color_manual(values = pal_class) + 
  scale_fill_manual(values = pal_class) + 
  theme_hpa(angled = T)


dat_healthy <- 
  healthy_importance |> 
  filter(class %in% example_diseases) |> 
  group_by(term, class) |> 
  summarise(abs_estimate = mean(abs(estimate))) |> 
  arrange(-abs_estimate)  |> 
  left_join(healthy_importance |>
              group_by(class) |> 
               count(term), by = c("term", "class")) |> 
   left_join(resource_meta
            |> distinct(Disease, Class), by = c("class" = "Disease")) |>
  mutate(Type = "Healthy") 

#p1 <- 
p1 <- 
  dat_healthy |> 
 
  ggplot(aes(abs_estimate, n, color = Class)) +
  geom_point() +
  geom_text_repel(data = filter(dat_healthy, abs_estimate > 0.4),
                  aes(label = term), size = 3) +
  facet_wrap(~class, scales = "free_x", nrow = 1) +
  scale_color_manual(values = pal_class) +
  theme_hpa() +
  ggtitle("Healthy")


dat_hmulti<- 
  multiclass_importance |> 
  filter(class %in% example_diseases) |> 
  group_by(term, class) |> 
  summarise(abs_estimate = mean(abs(estimate))) |> 
  arrange(-abs_estimate)  |> 
  left_join(multiclass_importance |>
              group_by(class) |> 
               count(term), by = c("term", "class")) |> 
   left_join(resource_meta
            |> distinct(Disease, Class), by = c("class" = "Disease")) |>
  mutate(Type = "Healthy") 

#p1 <- 
p2 <- 
  dat_hmulti |> 
 
  ggplot(aes(abs_estimate, n, color = Class)) +
  geom_point() +
  geom_text_repel(data = filter(dat_hmulti, abs_estimate > 0.4),
                  aes(label = term), size = 3) +
  facet_wrap(~class, scales = "free_x", nrow = 1) +
  scale_color_manual(values = pal_class) +
  theme_hpa() +
  ggtitle("Multiclass")


p1 / p2
ggsave(savepath("number_features-healthy_multi.png"), h = 8, w = 12)

healthy_importance |> 
  filter(class == "Acute myeloid leukemia") |> 
  filter(abs(estimate) > 0) |> 
  count(term) |> 
  arrange(-n)

```

### Example diseases

```{r}

example_diseases <- c("Rheumatoid arthritis",
                      "Acute venous thromboembolism",
                      "Schizophrenia",
                      "MASLD", 
                      "Malaria")

imp_plots <- 
  map(example_diseases, \(disease) {
    
    imp <- 
      multiclass_importance |> 
      filter(class == disease)
    
    current_class <- 
      resource_meta |> 
      filter(Disease == disease) |> 
      pull(Class) |> 
      unique()
    
    plot_top_proteins_seed(imp, 
                           plot_color = pal_class[[current_class]],
                           n = 10) +
      ggtitle(disease) 
  })

imp_plots_healthy <- 
  map(example_diseases, \(disease) {
    
    imp <- 
      healthy_importance |> 
      filter(class == disease)
    
    current_class <- 
      resource_meta |> 
      filter(Disease == disease) |> 
      pull(Class) |> 
      unique()
    
    plot_top_proteins_seed(imp, 
                           plot_color = pal_class[[current_class]],
                           n = 10) +
      ggtitle(disease)
    
    
  })

imp_plots[[1]] / imp_plots[[2]] / imp_plots[[3]] / imp_plots[[4]] / imp_plots[[5]] 
ggsave(savepath("features_selected_diseases.pdf"), h = 16, w = 4)


p1 <- 
  imp_plots[[1]] | imp_plots[[2]] | imp_plots[[3]]| imp_plots[[4]]| imp_plots[[5]] +
  ggtitle("Multiclass")

p2 <- 
  imp_plots_healthy[[1]] | imp_plots_healthy[[2]] | imp_plots_healthy[[3]]| imp_plots_healthy[[4]]| imp_plots_healthy[[5]] +
  ggtitle("Healthy")

ggsave(savepath("features_selected_diseases.pdf"), h = 4, w = 16)
ggsave(savepath("features_selected_diseases.pdf"), h = 4, w = 16)


p2/p1
ggsave(savepath("features_selected_diseases.png"), h = 8, w = 16)


```


### Compare

```{r}

dat_compare_imp <- 
  multiclass_importance |> 
  select(term, class, estimate, seed) |> 
  group_by(term, class) |> 
  summarise(abs_estimate_multiclass = mean(abs(estimate)),
            n_multiclass = n_distinct(seed)) |> 
  left_join(healthy_importance |> 
              select(term, class, estimate, seed) |> 
              group_by(term, class) |> 
              summarise(abs_estimate_healthy = mean(abs(estimate)),
                        n_healthy = n_distinct(seed)), by = c("term", "class")) |> 
  mutate(n_multiclass = ifelse(is.na(n_multiclass), 0, n_multiclass),
         n_healthy = ifelse(is.na(n_healthy), 0, n_healthy),
         abs_estimate_multiclass = ifelse(is.na(abs_estimate_multiclass), 0, abs_estimate_multiclass), 
         abs_estimate_healthy = ifelse(is.na(abs_estimate_healthy), 0, abs_estimate_healthy))

dat_compare_imp |> 
  filter(class %in% example_diseases) |> 
  left_join(resource_meta |> 
              distinct(Disease, Class), by = c("class" = "Disease")) |> 
  ggplot(aes(n_multiclass, n_healthy, color = Class)) +
  geom_point() +
  geom_text_repel(aes(label = term)) +
  facet_wrap(~class, nrow = 1) +
  scale_color_manual(values = pal_class) +
  theme_hpa()

ggsave(savepath("comparison_imp_n.png"), h = 7, w = 14)

dat_compare_imp |> 
  filter(class %in% example_diseases) |> 
  left_join(resource_meta |> 
              distinct(Disease, Class), by = c("class" = "Disease")) |> 
  ggplot(aes(abs_estimate_multiclass, abs_estimate_healthy, color = Class)) +
  geom_point() +
  geom_text_repel(aes(label = term)) +
  facet_wrap(~class, nrow = 1) +
  scale_color_manual(values = pal_class) +
  theme_hpa()

ggsave(savepath("comparison_imp_estimate.png"), h = 7, w = 14)
```

### Feature overlap 


### Protein pleiotropy

```{r}

multiclass_importance_avg <- 
  multiclass_importance |> 
  group_by(class, term) |> 
  summarise(avg_abs_estimate = mean(abs(estimate))) 
  
multiclass_importance_avg |> 
  count(class) |> 
  arrange(n) |> 
  left_join(resource_meta |> 
              distinct(Disease, Class), by = c("class" = "Disease")) |> 
  mutate(class = factor(class, levels = disease_class_order_ml)) |> 
  ggplot(aes(class, n, fill = Class)) +
  geom_col() +
  scale_fill_manual(values = pal_class) +
  theme_hpa(angled = T)

total_n_contribution <- 
  multiclass_importance_avg |> 
  ungroup() |> 
  count(term) |> 
  arrange(-n) |> 
  head(30)

multiclass_importance_avg |> 
  filter(term %in% total_n_contribution$term) |> 
  left_join(resource_meta |> 
              distinct(Disease, Class), by = c("class" = "Disease")) |> 
  group_by(term) |> 
  count(Class) |>
  left_join(total_n_contribution |> 
              rename(total_n = n), by = "term") |>
  ggplot(aes(fct_reorder(term, -total_n), n, fill = Class)) +
  geom_col() +
  scale_fill_manual(values = pal_class) +
  theme_hpa(angled = T) +
  xlab("")

```

### Network

```{r}
ml_features <- 
  multiclass_importance_avg |> 
  filter(avg_abs_estimate > 0.05) |>
  select(proteins = term,
         disease = class)


overlap_diseases <- 
  expand_grid(disease1 = unique(ml_features$disease), 
              disease2 = unique(ml_features$disease))   |> 
    filter(disease1 != disease2) |> 
    group_by_all() |> 
    do({
      
      d1 <- .$disease1
      d2 <- .$disease2
      
      disease1 <- 
        ml_features |> 
        filter(disease == d1) |> 
        pull(proteins)
      
      disease2 <- 
        ml_features |> 
        filter(disease == d2) |> 
        pull(proteins)
      
      ov <- length(intersect(disease1, disease2)) 
      
      tibble(disease1 = d1,
             disease2 = d2,
             overlap = ov)
      
    }) 

a <- 
  overlap_diseases |>
  group_by(disease1) |> 
  top_n(3, overlap) |> 
  ungroup() |> 
  filter(overlap > 50)

tg <- 
  as_tbl_graph(a, directed = FALSE) |> 
  activate(nodes) |> 
  left_join(resource_meta |> 
              select(name = Disease, Class) |> 
              distinct(), by = "name")

ggraph(tg, layout = "stress") +
  geom_edge_link(aes(width = overlap), color = "grey") +
  geom_node_point(aes(color = Class), size = 10) +
  geom_node_text(aes(label = name), repel = TRUE) +
  scale_color_manual(values = pal_class) +
  theme_void() 

ggsave(savepath("de_overlap.png"), h = 8, w = 8)
```


### ukb

```{r}
dat <- 
  multiclass_importance |> 
  group_by(term, class) |> 
  summarise(n = n_distinct(seed),
            importance = mean(abs(estimate))) |> 
  ungroup() |> 
  group_by(class) |> 
  mutate(rank_importance = rank(-importance),
         rank_n = rank(-n)) |> 
  ungroup() |> 
  mutate(sum_rank = rank_importance + rank_n) 

dat_filter <- 
  dat |> 
  filter(n > 40) #|> 
  #count(class) |> 
  #arrange(-n)

saveRDS(dat_filter, savepath_data("ML_test", "ml_multiclass_over50.rds"))


```

## Performances

Performances per disease against comparisons:

### Extract predictions

```{r}

class_predictions <- 
  map_df(ml_class, \(ml_res) {
    
   ml_res$predictions
    
  })

healthy_predictions <- 
  map_df(ml_healthy, \(ml_res) {
    
   ml_res$predictions |> 
      rename(class = Disease) |> 
      mutate(.pred_class = )
    
  })

multiclass_predictions <- ml_multiclass$predictions

```

### Calculate performance

```{r}

predictions <- 
  healthy_predictions |> 
  filter(!is.na(`.pred_0_Bipolar disorder`)#,
         #seed == 1
         ) 

predictions <- 
  class_predictions |> 
  filter(!is.na(`.pred_Abdominal aortic aneurysm`))

do_roc <- function(predictions) {
  
  
  clases <- 
    predictions |> 
    distinct(.pred_class) |> 
    pull()
  
  if(length(classes) > 2) {
    
    
    
  } else {
    
    roc <- 
      predictions |>
      roc_curve(truth = Disease, paste0(".pred_1_", disease), event_level = "second")
    
    # AUC
    res <- 
      predictions |> 
      rename(prediction = paste0(".pred_1_", disease)) |> 
      mutate(Disease = factor(Disease, levels = c("0_Healthy", paste0("1_", disease))))
    
    auc <- pROC::auc(res$Disease, res$prediction)
    ci <- pROC::ci.auc(res$Disease, res$prediction) 
    
    roc |> 
      mutate(AUC = auc,
             CI = ci)
    
  }
  
}

```


### Plot AUC comparison

```{r}
roc_class |> 
  distinct(Group, Seed, AUC) |> 
  left_join(meta_disease |> 
              distinct(Group = Disease, Class)) |> 
  mutate(Group = factor(Group, levels = disease_class_order)) |> 
  ggplot(aes(Group, AUC, color = Class)) +
  geom_quasirandom() +
  geom_boxplot(fill =NA, outlier.colour = NA) +
  scale_color_manual(values = pal_class) +
  theme_hpa(angled = T)


all_summary <- 
  roc_all |> 
  distinct(Group, Seed, AUC) |> 
  group_by(Group) |> 
  summarise(mean_auc = mean(AUC),
            sd_auc = sd(AUC)) |> 
  mutate(Type = "All diseases")

class_summary <- 
  roc_class |> 
  distinct(Group, Seed, AUC) |> 
  group_by(Group) |> 
  summarise(mean_auc = mean(AUC),
            sd_auc = sd(AUC)) |> 
  mutate(Type = "Class")


all_summary |> 
  bind_rows(class_summary)|> 
  bind_rows(roc_healthy |> 
  distinct(Group = Class, mean_auc = AUC, Type)) |> 
  left_join(meta_disease |> 
              distinct(Group = Disease, Class)) |> 
  mutate(Group = factor(Group, levels = disease_class_order)) |> 
  ggplot(aes(Group, mean_auc, color = Type)) +
  geom_point() +
  geom_segment(aes(x = Group, xend = Group, 
                   y = mean_auc - sd_auc, 
                   yend = mean_auc + sd_auc)) +
  scale_color_manual(values = c("#B39BC8","#D9B382","#C75D4D")) +
  facet_grid(~Class, scales = "free_x", space = "free") +
 # scale_color_manual(values = pal_class) +
  theme_hpa(angled = T)

ggsave(savepath_results("Manuscript-figures", "auc_comparison.pdf"), width = 12, height = 5)

"#B39BC8"
"#D9B382"
"#C75D4D"

```


### Plot ROC for selected diseases

### Plot confusion matrix

```{r}

predictions_all |> 
  mutate(Disease = factor(Disease),
         .pred_class = factor(.pred_class)) |>
  conf_mat(truth = Disease, estimate = .pred_class)  |> 
  plot_cm(percentage = T) 

# predictions_all |>
#   filter(Seed == 13) |>
#   mutate(Disease = factor(Disease),
#          .pred_class = factor(.pred_class)) |>
#   conf_mat(truth = Disease, estimate = .pred_class)  |>
#   plot_cm(percentage = T)

ggsave(savepath_results("Manuscript-figures", "confusion_matrix_multiclass.pdf"), width = 10, height = 10)
```


