---
title: "07_validation_cancer"
output: html_document
autor: "María Bueno Álvez"
date: "2024-12-13"
editor_options: 
  chunk_output_type: console
---

# Set up

```{r}
source("scripts/functions/functions_analyses.R")
source("scripts/functions/functions_utility.R")
source("scripts/functions/functions_visualization.R")
source("scripts/functions/themes_palettes.R")


# Read in data - UCAN
data_ucan <-
  read_tsv(
    "../Human-disease-blood-atlas/data/final_data/HPA/v24_2/ucan_data_ht_phase2.tsv"
  )
meta_ucan <-
  read_tsv(
    "../Human-disease-blood-atlas/data/final_data/HPA/v24_2/ucan_meta_ht_phase2.tsv"
  )

# Read in data - pan-disease
data_disease <-
  read_tsv(
    "../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_data_phase1.tsv"
  )
meta_disease <-
  read_tsv(
    "../Human-disease-blood-atlas/data/final_data/HPA/v24_2/disease_meta_phase1.tsv"
  )

# Other
olink_targets <- read_tsv("data/other/targets_olink_platforms.tsv")

cancers <- c(
  "Breast cancer",
  "Ovarian cancer",
  "Prostate cancer",
  "Colorectal cancer",
  "Lung cancer"
)
```



# Cohort overview

Explore number of samples and representative protein in the discovery U-CAN cohort (Olink Explore 1536) and the validation U-CAN cohort (Olink Explore HT).

## Number of samples 

```{r}
p1 <-
  meta_disease |>
  filter(Disease %in% cancers) |>
  count(Disease) |>
  mutate(Disease = factor(Disease, levels = rev(cancers))) |>
  ggplot(aes(Disease, n, fill = Disease)) +
  geom_col(show.legend = F) +
  geom_text(aes(label = n), vjust = 0.5) +
  coord_flip() +
  scale_y_reverse(limits = c(500, 0)) +
  scale_fill_manual(values = pal_ucan) +
  xlab("") +
  theme_hpa(axis_y = F)

p2 <-
  meta_ucan |>
  filter(DAid %in% data_ucan$DAid) |>
  count(Disease) |>
  mutate(Disease = factor(Disease, levels = rev(cancers))) |>
  ggplot(aes(Disease, n, fill = Disease)) +
  geom_col(show.legend = F) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 500)) +  # Ensure same scale
  geom_text(aes(label = n), vjust = 0.5) +
  theme_hpa() +
  scale_fill_manual(values = pal_ucan) +
  xlab("")

p1 + p2

ggsave(
  savepath_results("Manuscript-figures", "samples_ucan.pdf"),
  h = 3,
  w = 9
)
```
## Example marker - PAEP

```{r}
data_disease |>
  filter(Assay == "PAEP") |>
  left_join(meta_disease, by = "DAid") |>
  filter(Disease %in% cancers) |>
  mutate(Disease = factor(Disease, levels = cancers)) |>
  ggplot(aes(Disease, NPX, fill = Disease, color = Disease)) +
  geom_quasirandom(alpha = 0.6, show.legend = F) +
  geom_boxplot(
    alpha = 0.5,
    outlier.color = NA,
    color = "grey20",
    show.legend = F
  ) +
  scale_fill_manual(values = pal_ucan) +
  scale_color_manual(values = pal_ucan) +
  theme_hpa(axis_x_title = T) +
  xlab("Cancer type") +
  
  data_ucan |>
  filter(Assay == "PAEP") |>
  left_join(meta_ucan, by = "DAid") |>
  mutate(Disease = factor(Disease, levels = cancers)) |>
  ggplot(aes(Disease, NPX, fill = Disease, color = Disease)) +
  geom_quasirandom(alpha = 0.6, show.legend = F) +
  geom_boxplot(
    alpha = 0.5,
    outlier.color = NA,
    color = "grey20",
    show.legend = F
  ) +
  scale_fill_manual(values = pal_ucan) +
  scale_color_manual(values = pal_ucan) +
  theme_hpa(axis_x_title = T) +
  xlab("Cancer type")

ggsave(savepath_results("Manuscript-figures", "PAEP.pdf"),
       h = 3.5,
       w = 9)
```
# Differential expression

Compare differential expression in the discovery U-CAN cohort (Olink Explore 1536) and the validation U-CAN cohort (Olink Explore HT).

## Olink Explore HT

```{r}
female_diseases <- c("Ovarian cancer", "Breast cancer")
male_diseases <- c("Prostate cancer")

de_ucan_p2 <-
  map_df(cancers, function(cancer) {
    message("Processing disease: ", cancer)
    
    # Combine disease and control metadata
    disease_meta <-
      meta_ucan |>
      mutate(Disease = ifelse(Disease == cancer, cancer, "Control")) |>
      select(DAid = DAid, Disease, Sex, Age) |>
      mutate(BMI = NA)
    
    # Filter controls based on sex-specific diseases
    if (cancer %in% female_diseases) {
      disease_meta <-
        disease_meta |>
        filter(Sex == "F")
    } else if (cancer %in% male_diseases) {
      disease_meta <-
        disease_meta |>
        filter(Sex == "M")
    } else {
      disease_meta <-
        disease_meta
    }
    
    disease_data <-
      data_ucan |>
      filter(DAid %in% disease_meta$DAid) |>
      select(DAid, Assay, NPX) |>
      pivot_wider(names_from = "Assay", values_from = "NPX")
    
    disease_res <-
      do_limma_disease(
        data_wide = disease_data,
        metadata = disease_meta,
        disease = cancer,
        controls = "Cancers",
        cutoff = 0.5,
        correct = F
      )
  })
```

## Olink Explore 1463

```{r}
de_ucan_p1 <-
  map_df(cancers, function(cancer) {
    message("Processing disease: ", cancer)
    
    # Combine disease and control metadata
    disease_meta <-
      meta_disease |>
      filter(Disease %in% cancers) |>
      mutate(Disease = ifelse(Disease == cancer, cancer, "Control")) |>
      select(DAid = DAid, Disease, Sex, Age) |>
      mutate(BMI = NA)
    
    # Filter controls based on sex-specific diseases
    if (cancer %in% female_diseases) {
      disease_meta <-
        disease_meta |>
        filter(Sex == "F")
    } else if (cancer %in% male_diseases) {
      disease_meta <-
        disease_meta |>
        filter(Sex == "M")
    } else {
      disease_meta <-
        disease_meta
    }
    
    disease_data <-
      data_disease |>
      filter(DAid %in% disease_meta$DAid) |>
      select(DAid, Assay, NPX) |>
      pivot_wider(names_from = "Assay", values_from = "NPX")
    
    disease_res <-
      do_limma_disease(
        data_wide = disease_data,
        metadata = disease_meta,
        disease = cancer,
        controls = "Cancers",
        cutoff = 0.5,
        correct = F
      )
  })
```

## Volcano plots 

```{r}
data_volcanos <-
  map_df(cancers, function(cancer) {
    sig_ucan_p1 <-
      de_ucan_p1 |>
      filter(Disease == cancer) |>
      filter(sig != "not significant")
    
    de_ucan_p2 |>
      filter(Disease == cancer) |>
      left_join(olink_targets, by = "Assay") |>
      mutate(
        Type = case_when(
          Assay %in% sig_ucan_p1$Assay &
            sig != "not significant" ~ "Replicated from Olink Explore 1463",
          Platform != "Olink Explore 1463" &
            sig != "not significant" ~ "New from Olink Explore HT",
          T ~ "Other"
        )
      )
  })

labels <-
  data_volcanos |>
  mutate(Disease = factor(Disease, levels = rev(cancers))) |>
  group_by(Disease) |>
  top_n(n = 5, wt = -log10(adj.P.Val))


plot_volcanos <-
  data_volcanos |>
  mutate(Disease = factor(Disease, levels = rev(cancers))) |>
  ggplot(aes(
    x = logFC,
    y = -log10(adj.P.Val),
    color = Type,
    label = Assay
  )) +
  geom_point(size = 1, alpha = 0.4) +
  geom_text_repel(data = labels,
                  size = 2,
                  show.legend = F) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = 'dashed',
    color = "darkgrey"
  ) +
  geom_vline(
    xintercept = -0.5,
    linetype = 'dashed',
    color = "darkgrey"
  ) +
  geom_vline(xintercept = 0.5,
             linetype = 'dashed',
             color = "darkgrey") +
  scale_color_manual(values = pal_ucan_replication) +
  facet_wrap( ~ Disease, scales = "free", nrow = 1) +
  theme_hpa() +
  theme(
    axis.text = element_text(size = 8),
    legend.position = "top",
    strip.text.y = element_blank()
  ) +
  labs(y = expression(-log[10] ~ adjusted ~ italic(P) ~ value))

ggsave(savepath_results("Manuscript-figures", "UCAN_DE.pdf"),
       h = 3.5,
       w = 12)
```

## Save data

```{r}
comparison_data <-
  de_ucan_p1 |>
  select(Assay, logFC_p1 = logFC, adj_pval_p1 = adj.P.Val, Disease) |>
  left_join(
    de_ucan_p2 |>
      select(Assay, logFC_p2 = logFC, adj_pval_p2 = adj.P.Val, Disease),
    by = c("Assay", "Disease")
  )

comparison_data |>
  rename(
    `logFC Olink Explore 1563` = logFC_p1,
    `p.adjusted Olink Explore 1563` = adj_pval_p1,
    `logFC Olink Explore HT` = logFC_p2,
    `p.adjusted Olink Explore HT` = adj_pval_p2
  ) |>
  relocate(Disease) |>
  write_csv(savepath_data("Supplementary-files", "Data_S8_ucan_de.csv"))
```







# Machine learning

Train machine learning models in the discovery U-CAN cohort (Olink Explore 1536) using 10-fold CV and apply to validation U-CAN cohort (Olink Explore HT).

## Phase 1


```{r}
# Lists of diseases
female_diseases <- c("Ovarian cancer", "Breast cancer")
male_diseases <- c("Prostate cancer")

# Prepare data
ucan_p1_wide <-
  data_disease |>
  select(DAid, Assay, NPX) |>
  pivot_wider(names_from = Assay, values_from = NPX) |>
  left_join(meta_disease |>
              select(DAid, Disease), by = "DAid") |>
  filter(Disease %in% cancers)

models_ucan_p1 <-
  map(cancers, function(cancer) {
    cat(paste0("\nPreparing training and testing data for ", cancer))
    
    ucan_data <-
      ucan_p1_wide |>
      rename(Variable = Disease) |>
      mutate(Variable = case_when(
        Variable == cancer ~ paste0("1_", Variable),
        Variable != cancer ~ paste0("0_Control")
      )) |>
      mutate(Variable = factor(Variable))
    
    
    if (cancer %in% female_diseases) {
      ucan_data <-
        ucan_data |>
        left_join(meta_disease |>
                    select(DAid, Sex), by = "DAid") |>
        filter(Sex == "F") |>
        select(-Sex)
    } else if (cancer %in% male_diseases) {
      ucan_data <-
        ucan_data |>
        left_join(meta_disease |>
                    select(DAid, Sex), by = "DAid") |>
        filter(Sex == "M") |>
        select(-Sex)
    } else {
      ucan_data <- ucan_data
    }
    
    
    # Recipe with ML steps
    discrete_recipe <-
      recipe(Variable ~ ., data = ucan_data) |>
      update_role(DAid, new_role = "id") |>
      step_normalize(all_numeric()) |>
      step_nzv(all_numeric()) |>
      step_impute_knn(all_numeric())
    
    # LASSO model specifications
    glmnet_specs <-
      logistic_reg() |>
      set_mode("classification") |>
      set_engine("glmnet") |>
      set_args(penalty = tune(),
               mixture = 1)
    
    # ML workflow
    glmnet_wflow <-
      workflow() |>
      add_recipe(discrete_recipe) |>
      add_model(glmnet_specs)
    
    # Define glmnet grid
    set.seed(213)
    glmnet_grid <-
      glmnet_wflow |>
      extract_parameter_set_dials() |>
      grid_latin_hypercube(size = 20)
    
    # Define the resamples (CV)
    set.seed(213)
    ml_rs <- vfold_cv(ucan_data, v = 10, strata = Variable)
    
    # Define the evaluation metrics
    eval_metrics <- metric_set(roc_auc)
    
    # Define control_grid
    set.seed(213)
    ctrl <-
      control_grid(
        save_pred = TRUE,
        parallel_over = "everything",
        event_level = "second"
      )
    
    cat(paste0("\nFitting glmnet model for ", cancer))
    
    # Glmnet grid search
    set.seed(213)
    glmnet_res <-
      glmnet_wflow |>
      tune_grid(
        resamples = ml_rs,
        grid = glmnet_grid,
        control = ctrl,
        metrics = eval_metrics
      )
    
    cat(paste0("\nSelecting best performing model for ", cancer))
    
    metrics_train <-
      glmnet_res |>
      collect_metrics()
    
    predictions_train <-
      glmnet_res |>
      collect_predictions()
    
    roc_train <-
      glmnet_res |>
      collect_predictions() |>
      group_by(id) |>
      roc_curve(truth = Variable,
                colnames(predictions_train)[[2]],
                event_level = "second")
    
    combined_roc_auc <-
      map_df(c(1:10), function(fold) {
        if (fold < 10) {
          fold <- paste0(0, fold)
        } else {
          fold <- fold
        }
        
        res <-
          predictions_train |>
          filter(id == paste0("Fold", fold)) |>
          rename(prediction = paste0(".pred_1_", cancer)) |>
          mutate(Variable = factor(Variable, levels = c("0_Control", paste0("1_", cancer))))
        
        auc <- pROC::auc(res$Variable, res$prediction)
        ci <- pROC::ci.auc(res$Variable, res$prediction)
        
        roc_train |>
          filter(id == paste0("Fold", fold)) |>
          mutate(
            AUC = as.numeric(auc),
            CI_lower = as.numeric(ci)[[1]],
            CI_upper = as.numeric(ci)[[3]]
          )
        
      })
    
    
    # Select best hyperparameter
    best_glmnet <-
      select_best(glmnet_res, metric = "roc_auc") |>
      select(-.config)
    
    #Finalize the workflow and fit the final model
    glmnet_wflow <-
      glmnet_wflow |>
      finalize_workflow(best_glmnet)
    
    # Final fit
    final_glmnet_fit <- fit(glmnet_wflow, data = ucan_data)
    
    # Extract proteins
    protein_importance <-
      final_glmnet_fit |>
      extract_fit_parsnip() |>
      tidy() |>
      filter(estimate != 0,
             term != "intercept")
    
    return(
      list(
        disease = cancer,
        model = final_glmnet_fit,
        fold_roc = combined_roc_auc,
        predictions_train = predictions_train,
        protein_importance = protein_importance
      )
    )
    
  }) |>
  set_names(cancers)


saveRDS(models_ucan_p1,
        savepath_data("ML", "models_ucan_p1_ext.rds"))
```

## Phase 2

```{r}
# Prepare data
ucan_p2_wide <-
  data_ucan |>
  select(DAid, Assay, NPX) |>
  pivot_wider(names_from = Assay, values_from = NPX) |>
  left_join(meta_ucan |>
              select(DAid, Disease), by = "DAid")

# Deal with missing features
missing_ht <-
  c(
    'ADAMTS16',
    'NTproBNP',
    'CD209',
    'MB',
    'LILRB2',
    'PLAT',
    'LILRB1',
    'AMY2A',
    'AMY2B',
    'DEFA1_DEFA1B',
    'FCGR3B',
    'KIR2DL3',
    'UXS1',
    'SFTPA1',
    'LGALS7_LGALS7B',
    'CXCL12',
    'SELPLG',
    'PTPRM',
    'LTA',
    'MICB_MICA',
    'RAB6A',
    'BID',
    'PCDH1',
    'DDX58',
    'CLSTN1',
    'CDHR1',
    'ULBP2',
    'NGF',
    'ARSA',
    'CGA',
    'WARS'
  )

ucan_p2_wide[missing_ht] <- NA

# Run models for UCAN P2
results_ucan_p2 <-
  map(cancers, function(cancer) {
    current_data <-
      ucan_p2_wide |>
      mutate(Variable = case_when(
        Disease == cancer ~ paste0("1_", cancer),
        Disease != cancer ~ paste0("0_Control")
      )) |>
      select(-Disease) |>
      mutate(Variable = factor(Variable))
    
    current_model <- models_ucan_p1[[cancer]]$model
    new_predictions <-
      predict(current_model, new_data = current_data, type = "prob")
    
    new_roc <-
      new_predictions |>
      mutate(Variable = current_data$Variable) |>
      roc_curve(truth = Variable,
                paste0(".pred_1_", cancer),
                event_level = "second")
    
    # AUC
    new_res <-
      new_predictions |>
      mutate(Variable = current_data$Variable) |>
      rename(prediction = paste0(".pred_1_", cancer)) |>
      mutate(Variable = factor(Variable, levels = c("0_Control", paste0("1_", cancer))),
             DAid = current_data$DAid)
    
    auc <- pROC::auc(new_res$Variable, new_res$prediction)
    ci <- pROC::ci.auc(new_res$Variable, new_res$prediction)
    
    # Combine
    combined_roc_auc <-
      new_roc |>
      mutate(
        AUC = as.numeric(auc),
        CI_lower = as.numeric(ci)[[1]],
        CI_upper = as.numeric(ci)[[3]]
      )
    
    return(list(predictions = new_res,
                roc = combined_roc_auc))
  }) |>
  set_names(cancers)


saveRDS(results_ucan_p2, savepath_data("ML", "results_ucan_p2.rds"))
```

## Plot AUC

```{r}
roc_all <-
  map_df(cancers, function(cancer) {
    current_model_p1 <- models_ucan_p1[[cancer]]
    current_model_p2 <- results_ucan_p2[[cancer]]
    
    current_roc <-
      current_model_p1$fold_roc |>
      mutate(Cancer = cancer) |>
      mutate(Platform = "Olink Explore 1563") |>
      bind_rows(
        current_model_p2$roc |>
          mutate(Cancer = cancer) |>
          mutate(Platform = "Olink Explore HT",
                 id = "Validation")
      )
  })

tiles <-
  roc_all |>
  distinct(id, Cancer, Platform, AUC) |>
  mutate(Group = ifelse(id == "Validation", "Validation", "Training")) |>
  group_by(Cancer, Platform, Group) |>
  summarise(AUC = mean(AUC)) |>
  select(Cancer, Platform, AUC) |>
  mutate(
    specificity = 0.2,
    sensitivity = ifelse(Platform == "Olink Explore 1563", 0.1, 0.2),
    Cancer = factor(Cancer, levels = rev(cancers))
  ) |>
  mutate(Cancer = factor(Cancer, levels = rev(cancers)))

roc_all |>
  mutate(Cancer = factor(Cancer, levels = rev(cancers))) |>
  ggplot(aes(
    x = 1 - specificity,
    y = sensitivity,
    color = Platform,
    group = id
  )) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "grey50",
    size = 0.8
  ) +  
  geom_line(alpha = 0.4, size = 0.6) +  
  geom_tile(
    data = tiles,
    aes(
      x = 1 - specificity,
      y = sensitivity,
      fill = Platform
    ),
    width = 0.2,
    height = 0.1,
    show.legend = FALSE,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = tiles,
    aes(
      x = 1 - specificity,
      y = sensitivity,
      label = round(AUC, 2)
    ),
    size = 3,
    color = "white",
    fontface = "bold",
    show.legend = FALSE,
    inherit.aes = FALSE
  ) +
  facet_wrap( ~ Cancer, nrow = 1, scales = "free") +  
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 1)) +
  scale_color_manual(values = pal_platforms) +
  scale_fill_manual(values = pal_platforms) +
  theme_hpa(facet_title = FALSE) +
  theme(
    axis.text = element_text(size = 8),
    legend.position = "top",
    strip.text.y = element_blank()
  )

ggsave(savepath_results("Manuscript-figures", "UCAN_ML.pdf"),
       h = 3.5,
       w = 10)
```

## Save performance

```{r}
roc_all |>
  distinct(id, AUC, CI_lower, CI_upper, Cancer, Platform) |>
  relocate(Platform, Cancer) |>
  mutate(Cancer = factor(Cancer), levels = cancers) |>
  arrange(Cancer) |>
  write_csv(savepath_data("Supplementary-files", "Data_S9_UCAN_ML_performance.csv"))
```
